---
title: "Journals flipped by Springer Nature between 2015 and 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  fig.width = 9,
  fig.height = 6
)
options(scipen = 999, digits = 2)
knitr::knit_hooks$set(inline = function(x) {
      if(is.numeric(x)){
          return(prettyNum(x, big.mark=","))
      }else{
          return(x)
       }
   })
```

The aim of this document is to examine how many journals published by Springer Nature changed their business model between 2015 - 2018.

## Methods and Data

[Bernd-Christoph Kämper](https://twitter.com/bckaemper/status/993477985327108101) from Stuttgart University Library pointed me to [this catalogue of journals offered by Springer Nature to librarians](https://www.springernature.com/gp/librarians/licensing/journals-price-list). Let's obtain the journal  spreadsheets for 2015 and 2018, and load them into R.

```{r}
library(tidyverse)
library(rio)
library(viridis)
sn_2018 <-
  rio::import("https://resource-cms.springernature.com/springer-cms/rest/v1/content/15226614/data/v7", skip = 8) %>%
  as_data_frame() %>%
  # add year column
  mutate(file_year = 2018)
sn_2015 <- rio::import("https://resource-cms.springernature.com/springer-cms/rest/v1/content/10619914/data/v1", skip = 8) %>%
  as_data_frame() %>%
  # add year column
  mutate(file_year = 2015)
# let's bind them
bind_rows(sn_2018, sn_2015) -> my_data
my_data
```

To get flipped journals, we will examine which journals changed their open access business model.

```{r}
my_data %>%
  distinct(`Open Access Option`, Title) %>%
  group_by(Title) %>%
  filter(!is.na(`Open Access Option`)) %>%
  filter(n() > 1) -> flipped_jns
```

## Results

The following figure displays the number of journals and their underlying business model.


```{r}
my_data %>%
  count(file_year, `Open Access Option`) %>% 
  # remove NAs
  mutate(`Open Access Option` = ifelse(is.na(`Open Access Option`), "not reported", `Open Access Option`)) %>%
  ggplot(aes(factor(file_year), n, fill = `Open Access Option`)) + 
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_x_discrete(drop = FALSE) +
  viridis::scale_fill_viridis("OA Option", discrete = TRUE) +
  scale_y_continuous(labels=function(x) format(x, big.mark = " ", scientific = FALSE),
                       breaks= scales::pretty_breaks()) +
  # mimicking github.com/hrbrmstr/hrbrthemes 
  theme_minimal(base_family="Arial Narrow", base_size = 16) +
  theme(plot.margin=margin(30,30,30,30)) +
  theme(panel.grid.minor=element_blank()) +
  theme(axis.ticks=element_blank()) +
  theme(panel.grid.major.y=element_blank()) +
  theme(panel.border=element_blank()) +
  xlab("Year") +
  ylab("Number of Springer Nature journals")
```

Apparently, both the total number of journals and the number of hybrid open access journals increased between 2015 - 2018. Hybrid open access is still the dominant business model for Springer Nature journals.

We also asked about journals were the business model changed over the year. The following table summarizes the open access status for these flipped journals in 2018:

```{r}
sn_2018 %>%
  filter(Title %in% flipped_jns$Title) %>%
  count(`Open Access Option`) %>%
  knitr::kable()
```

Surprisingly, more journals changed the business model to Hybrid (Open Choice) than to fully open access, including journals like [Health Information Science Systems](https://link.springer.com/journal/13755), which was published under the brand "BioMed Central" in 2015, and is now published as toll-access journal.

Here's the list of all flipped journals with their most current open access status:

```{r}
sn_2018 %>%
  filter(Title %in% flipped_jns$Title) %>%
  select(Title, `Open Access Option`) %>%
  knitr::kable()
```

## Discussion and conclusion

Despite recent efforts to transition journals from a subscription-based model to fully open access, our findings suggest that hybrid open access is still prevalent for journals published by Springer Nature. In fact, more journals are flipped from a fully open access to hybrid open access, raising important questions about how to sustain fully open access journals.

## Session info

```{r}
sessionInfo()
```
