---
title: "Report Open Access status JASIST"
author: "Anne Hein"
output:
  html_document:
    keep_md: true
bibliography: crossref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      cache = TRUE, 
                      warning = FALSE, 
                      error = FALSE,
                      message = FALSE
                      )
```

## Introduction

@Piwowar2017 investigated the Open Access (OA) status of scholarly literature. They examined three samples of 100 000 articles each chosen from three populations, namely journal articles assigned a Crossref DOI, recent journal articles indexed in Web of Science, and articles viewed by users of Unpaywall. They estimate that at least 28 % of all scientific publications is OA, for the year 2015 they even obtain that almost half (45%) of the articles is OA.

In this brief report I investigated all articles with Crossref DOI published in the [Journal of the Association for Information Science and Technology (JASIST)](http://onlinelibrary.wiley.com/journal/10.1002/(ISSN)2330-1643) in the years from 2014 to 2017 with respect to their OA status.

To this end, I first imported the metadata of the articles published in JASIST between 2014 and 2017 using the rcrossref package by @CrossrefPack and determined the total number of publications in each of the four years. In a second step I used the roadoi package by @RoadoiPack to examine which of the articles are OA.

## Gathering bibliographic metadata

In this first step I used the `rcrossref::cr_works()` function to find the metadata of all articles published in JASIST (ISSN-Number: 2330-1635) which have a publication date between 1st of January 2014 and the 31st of December 2017. 

```{r cache = TRUE}
library(rcrossref)
metadata <- rcrossref::cr_works(
  filter = c(
    from_pub_date = "2014-01-01",
    until_pub_date = "2017-12-31",
    issn = "2330-1635"
    ),
  limit = 1000
)
metadata
```

I determined the number of articles published in each year. The results are displayed in the following figure.

```{r, fig.cap="Number of publications per year"}
library(dplyr)
library(ggplot2)
metadata$data %>%
  mutate(issued, issued = lubridate::parse_date_time(issued, c('y', 'ymd', 'ym'))) %>%
  mutate(issued, issued = lubridate::year(issued)) %>%
  group_by(issued) %>%
  summarize(pubs = n())  %>%
  # arrange(desc(pubs))
  ggplot(aes(issued, pubs)) + geom_bar(stat = "identity") + labs(title = "Number of publications per year", x  ="Publication year", y = "Number of publications")
```

## Availability of Open Access full text versions

In a second step, I applied the `roadoi::oadoi_fetch()` function to examine for which of the dois an Open Access full text version can be found by the oaDOI search engine.

```{r cache = TRUE}
library(roadoi)
oa_df <- roadoi::oadoi_fetch(dois = metadata$data$DOI, 
                             email = "anne.hein@mathematik.uni-goettingen.de")
oa_df
```

The following figure displays the number of articles that are OA and not OA, resp.

```{r}
oa_df %>%
  group_by(is_oa, year) %>%
  summarize(oa = n()) %>%
  ggplot(aes(year, oa, fill = is_oa)) + 
  geom_bar(stat = "identity") + 
  labs(title = "How many JASIST articles are openly available?", 
       subtitle = "Data from Crossref and oaDOI",
       x = "Publication year", 
       y = "Number of Publications") +
  scale_fill_manual(name = "OA status", values = c("#545E75", "#63ADF2") , labels = c("Non Open Access", "Open Access")) +
  theme_minimal(base_family="Arial Narrow")
```

Apparently, for the JASIST the percentage of articles that are OA is much lower than the estimated overall percentage of OA articles from @Piwowar2017. Furthermore, the percentage of OA articles decreased, especially for the last year.

## Acknowledgement

I thank Birgit Schmidt for offering me the internship during which I wrote this report and I especially thank Najko Jahn for showing me how to use R Markdown and for revising my work.

## Session Info

```{r, include=TRUE}
sessionInfo()
```


## References

